Diversification of Australian blind snakes

### 2022-11-15 manuscript.Rmd
updated results

## Phylogenetic hypothesis and taxon sampling
<!-- Based on our phylogenetic tree -->
We employ the recent time-calibrated phylogenomic tree of Australian blindsnake with outgroups. This tree includes 38 operational taxonomic units (OTUs) representing  __taxa representing __ described species of _Anilios_. However, because our phylogenetic data suggest some cryptic species, we adjust our taxon sampling fraction to XX/XX species. The genetically distinct OTUs that are not fully recognised species include 3 '_Anilios grypus_' and 2 '_Anilios ligatus_'.

@@

We used the R package `BioGeoBEARS` 1.1.1 [@matzkeBioGeoBEARS2013] to estimate the ancestral biome of blindsnakes in Australia. BioGeoBEARS provides a flexible likelihood-based framework that models how the branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events including processes like dispersal, extinction, or switching to adjacent biomes) or at the time of branching (cladogenetic events involving sympatric, vicariant and founder-event speciation). The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range each species are found, and a historical biogeographic model. We divided Australo-Papua into nine discrete biomes and assigned each Australian species to specific biome(s) based on their occurence records as described above. Biome classification were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World dataset [@olsonTerrestrial2001] and followed the nomenclature of previous Austrlia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021]. 
The phylogeny and distribution information were then applied to widely used historical biogeographical methods including Dispersal-Vicariance Analysis (DIVA; [@ronquistDispersalVicariance1997]), and likelihood implementations of Dispersal-Extinction-Cladogenesis (DEC; [@reeMaximum2008]) or DECLIKE, and BayArea [@landisBayesian2013] or BAYAREALIKE [@matzkeBioGeoBEARS2013]. We included two additional free parameters. The first parameter *j* assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter _x_ modifies the probability of dispersal based on distance between biomes [@vandamEvaluating2016] (see Appendix-method-biogeobears). We used likelihood ratio tests to determine if the additional free parameter is a justified improvement over the simpler base models, and selected the best overall model based on corrected Akaike Information Criterion (AICc) scores.

## Reconstruction of Australian paleo climate and environment-dependent diversification analyses
<!-- Consult Alex Skeels -->
Three paleographic digital elevation models (DEMs) providing paleotopography at resolution of 1&deg; x 1&deg; and 5 myr resolution  were obtained from @scoteseAtlas2021, @straumeGlobal2020, [VALDES]. We projected the paleoenvironmental reconstructions into an equal-area Behrmann projection of 110km x 110km following methods described in @hagenGen3sis2021. From the projected data we extracted temperature and aridity values across all grid cells that fall within Sahul by manually drawing a polygon that encompassed the continent from 40 mya to the present-day. Because reptiles are sensitive to extremely low and high temperature, we decided to to summarise the reconstructed paleo temperature data into three different data sets: minimum temperature, maximum temperature, and mean temperature.

To test whether and how diversification rates correlated with climate and aridity through time, we fitted several time- and environemnt-dependent likelihood diversification models using the `fit_bd` and `fit_env` function in the R package `RPANDA` [@morlonRPANDA2016] to the _Anilios_ phylogeny. Specifically, we fitted different combinations of pure-birth and birth-death models with speciation and/or extinction rates being constant or otherwise varying linearly or exponentially with the following environment variables: 1 mean global paleo temperature, 3 Australian paleo temperature, and 3 Australian paleo aridity. We accounted for missing taxa using the fractioning function (f = XX) in RPANDA. We then assessed which diversification model best fit the data based on AICc scores.

To visualise the dynamics of the radiation, we use ClaDS [@malietFast2022] implemented in `julia` to estimate speciation rates at the origin of each branch (hencforth branch-specific rates). ClaDS is designed to capture frequent shifts with small effects. At each speciation event, daughter lineages are assigned its own speciation rate sampled from a distribution parameterised on the parental rate [@malietFast2022].

@@

We further investigate whether broad patterns of diversification are associated with geographic range. Using the R package `hisse` v. 2.1.9 [@beaulieuDetecting2016], we fitted five biogeographic-dependent speciation-extinction models from the GeoSSE [@goldbergPhylogenetic2011] and GeoHiSSE [@caetanoHidden2018] framework to the tree and geographic range data. 
Since state speciation-extinction (SSE) models only allow three states, we restricted our geographic states to arid, non-arid, and widespread. Consistent with our BioGeoBEARS anaylsis, we classified species that primarily occur in the arid zone as arid (n = 8), species that occur primarily outside of the arid zone as non-arid (n = 24), and those that occupy both the arid and non-arid zone as widespread (n = 5). We assumed uniform taxon sampling for 89% in arid areas, 71% in non-arid areas, and 83% in widespread areas based on number of species in the phylogeny and all currently described species of _Anilios_.
The five models vary by whether they allow diversification rates to be associated with geographic range and/or with unobserved hidden states. Model 1 assumes equal diversification rates regardless of geographic range, model 2 allows geographic range to affect diversification rates (canonical GeoSSE model). Model 3 and 4 are derivation of models 1 and 2 but with unobserved hidden states (see Table GeoHiSSE table). The last model is a complementary model that removes cladogenetic effect entirely. We compared the AIC scores to select the best supported models, then averaged the estimates of diversification parameters and ancestral geographic range estimate across all models according to their AIC weights following @caetanoHidden2018.

## Niche evolution and spatial data
We reconstructed ancestral niche from present-only data using NEMo package. 
from the online database [Atlas of the Living Australia](doi.ala.org.au/doi/10.26197/ala.d92678b1-ad2d-437b-9457-9f52737ba003) (see Appendix-method-occurence).

## Statistical analyses
All statistical analyses were run in R version 4.2 (CITE R). We visualised diversification data using `ggplot2`, made map with `QGIS`.

@@

## Biogeographic history
We reconstructed the tropical grassland as the ancestral biome of Australian _Anilios_ (\@ref(fig:fig-biogeobears). The preferred biogeographic model is the constrained-dispersal matrix, DEC+j+x. 
Likelihood ratio test (LRT) indicates that founder event (parameter j) and distance matrix (free parameter _x_) in the more complex model provides statistically significant improvement over the simpler DEC, DEC+*j*, and DEC+*x* models (X^2^, DF = 2, D-statistic = 26.7, p-value < 0.001). 
Our result also supports the hypothesis that the ancestor of the Australian clade invaded the continent, but exactly where they lived prior to this invasion is equivocal between the Solomon Islands and New Guinea. 
After arriving to Australia in the late Oligocene - early Miocene (~25 MYA), the ancestral _Anilios_ initially diversified rapidly within less than 1 million years. 
- Our analysis is consistent with @marinTracing2013 conclusion that Australian blindsnakes, like several other Australian squamates, originated from tropical-mesic origins and expanded into the arid zone since the mid-Miocene [@chappleParallel2004;@melvilleEvolutionary2011].

## Biome transitions and current distributions
Following the hypothesized colonisation event from a single common ancestor, the diversifcation of _Anilios_ occured very rapidly. Descendents of the earliest split remained in the Tropical Grassland biome but within a million years later another split occured. 
The earliest biome transition was from tropical grassland into Temperate Forest around 25 MYA (95% HPD: 28 – 22 MYA) and the Arid zone 24 MYA (95% HPD: 28 – 21 MYA). 
Subsequent splits show transitions from the arid zone into the Mediterranean biome starting after 15 MYA. 
Current distributions of _Anilios_ are mostly concentrated in the tropical grassland biome closely followed by the arid zone and Mediterranean biome. 
Closely related species tend to share the same biome and occur around the same geographical areas.

- Range expansion was important for diversification ...
- Topography dyanmics [@brazModelling2021]
- Solomon Island 40 MYA [@hallCenozoic2002]

Recent biogeographic reconstruction recovered India as the origin for superfamily Typhlopoidea (Gerrhopilidae, Xenotyphlopidae and Typhlopidae) [@kleinEvolution2021].

@@

Paleoclimate reconstructions of aridity differed slightly between the two methods \@ref(fig:fig-paleoclimate-compare)a. Both aridity index shows increase in aridity post 10 mya, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021]. Temperature for Australia differed between the three methods \@ref(fig:fig-paleoclimate-compare)b. Reconstructed temperature values from Australia were higher than global temperature and stayed relatively constant compared to global temperature that shows slight decrease.

## Diversification
The best fitting model among constant, time-dependent, temperature-dependent, and aridity dependent models is 
This model indicate that X associate with speciation (alpha = ) and .....

Fit with Australia specific mean temperature receives better support than model with global temperature and exponential dependency of speciation rate with time.

@@

## Ancestral biome reconstruction
- Our result — that the ancestor of Anilios occupied tropical grasslands – is consistent with the result presented in @marinTracing2013. 
- We found transitions from XXX to XXX to be common as have been shown / as opposed to other Australian squamates (CITATIONS)
- Arid zone was not arid until ~20 Million Years ago. Ancestors in deeper time probably occupied temperate or tropical areas that subsequently became the arid zone we know see today. 
- The Australo-Papuan clade seems to have stayed in Australia for the most part but we were not able to include New Guinean endemic _Anilios erycinus_, the only recognised species of _Anilios_ not found in Australia. Therefore, we cannot make any conclusions about transitions between Australia and New Guinea even though at least three species (_A. torresianus_, _A. leucoproctus_, ...) are found in Northern Australia Torres Strait Islands, and southern parts of New Guinea.

>>>>

## Phylogenetic hypothesis and taxon sampling
<!-- Based on our phylogenetic tree -->
We employ the recent time-calibrated phylogenomic tree of Australian blindsnake with outgroups. This tree includes 38 operational taxonomic units (OTUs) representing  __taxa representing __ described species of _Anilios_. However, because our phylogenetic data suggest some cryptic species, we adjust our taxon sampling fraction to XX/XX species. The genetically distinct OTUs that are not fully recognised species include 3 '_Anilios grypus_' (2 the west and 1 in the east) and 2 '_Anilios ligatus_' (one from the east and one from the north). For ancestral range estimation analyses we also included _Ramphotyphlops multilineatus_, _Sundatyphlops polygrammicus_, and _Acutotyphlops subocularis_ as outgroup taxa.

@@

To estimate the ancestral biome of blind snakes in Australia, we used the R package `BioGeoBEARS` 1.1.1 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible likelihood-based framework that models how the branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events including processes like dispersal, extinction, or switching to adjacent biomes) or at the time of branching (cladogenetic events involving sympatric, vicariant and founder-event speciation). The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range each species are found, and a historical biogeographic model. We divided Australo-Papua into nine discrete biomes and assigned each Australian species to specific biome(s) based on their occurrence records as described above. Biome classification were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World dataset [@olsonTerrestrial2001] and followed the nomenclature of published Austrlia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021]. 
The phylogeny and distribution information were then applied to widely used historical biogeographical methods including likelihood implementations of Dispersal-Vicariance Analysis [@ronquistDispersalVicariance1997] (DIVALIKE), Dispersal-Extinction-Cladogenesis [@reeMaximum2008] (DECLIKE), and BayArea [@landisBayesian2013] (BAYAREALIKE) in `BioGeoBEARS` [@matzkeBioGeoBEARS2013]. We included two additional free parameters that we thought is relevant. The first parameter *j* assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter _x_ modifies the probability of dispersal based on distance between biomes [@vandamEvaluating2016] (see Appendix-method-biogeobears). We used likelihood ratio tests to determine if the additional free parameter is a justified improvement over the simpler base models, and selected the model that fitted the data best based on corrected Akaike Information Criterion (AICc) scores.

## Reconstruction of Australian paleo climate
<!-- Consult Alex Skeels -->
To estimate paleoclimate, we used three paleographic digital elevation models (DEMs) from @scoteseAtlas2021, @straumeGlobal2020, and [VALDES]. Each DEM provide paleotopography at resolution of 1&deg; x 1&deg; and 5 myr resolution. We projected the paleoenvironmental reconstructions into an equal-area Behrmann projection of 110km x 110km following methods described in @hagenGen3sis2021. From the projected data we extracted temperature and aridity values across all grid cells that fall within Sahul by manually drawing a polygon that encompassed the continent from 40 mya to the present-day. Because reptiles are sensitive to extremely low and high temperature, we decided to to summarise the reconstructed paleo temperature data into three different data sets: minimum temperature, maximum temperature, and mean temperature.

--More information about summarising values and determining aridity index ---

## Time ane environment-dependent diversification analyses

To test whether and how diversification rates correlated with climate and aridity through time, we fitted several time- and environemnt-dependent likelihood diversification models using the `fit_bd` and `fit_env` function in the R package `RPANDA` [@morlonRPANDA2016] to the _Anilios_ phylogeny. Specifically, we fitted different combinations of pure-birth and birth-death models with speciation and/or extinction rates being constant or otherwise varying linearly or exponentially with paleo environment as described above. We also fit the model to widely use mean global paleo temperature data set (CITE). We accounted for missing taxa using the fractioning function (f = XX) in RPANDA. We then assessed which diversification model best fit the data based on AICc scores.

To visualise the dynamics of the radiation, we use ClaDS [@malietFast2022] implemented in `julia` to estimate speciation rates at the origin of each branch (hencforth branch-specific rates) and at the tip. ClaDS is designed to capture frequent shifts in speciation rates with small effects. At each speciation event, daughter lineages are assigned its own speciation rate sampled from a distribution parameterised on the parental rate [@malietFast2022].

@@

We further investigated whether broad patterns of diversification are associated with geographic range. Using the R package `hisse` v. 2.1.9 [@beaulieuDetecting2016], we fitted five biogeographic-dependent speciation-extinction (SSE) models under the GeoSSE [@goldbergPhylogenetic2011] and GeoHiSSE [@caetanoHidden2018] framework to the phylogeny and geographic range data. 
Since SSE models only allow three states, we restricted our geographic states to species that have distributions in arid areas, non-arid areas, and those that are widespread across both arid and non-arid areas. 
Consistent with our BioGeoBEARS anaylsis, we classified species that primarily occur in the arid zone as arid (n = 8), species that occur primarily outside of the arid zone as non-arid (n = 24), and those that occupy both the arid and non-arid zone as widespread (n = 5). We assumed uniform taxon sampling for 89% in arid areas, 71% in non-arid areas, and 83% in widespread areas based on number of species in the phylogeny and all currently described species of _Anilios_.
The five SSE models vary by whether they allow diversification rates to be associated with geographic range and/or with unobserved 'hidden states'. Model 1 assumes equal diversification rates regardless of geographic range, model 2 allows geographic range to affect diversification rates (canonical GeoSSE model). Model 3 and 4 are derivation of models 1 and 2 but with unobserved hidden states (see Table GeoHiSSE table). The last model is a complementary model that removes cladogenetic effect entirely. We compared the AIC scores to select the best models. Following the approach in @caetanoHidden2018, we averaged the estimates of diversification parameters and ancestral geographic range estimate across all models according to their AIC weights.

<!-- ## Niche evolution and spatial data -->
<!-- We reconstructed ancestral niche from present-only data using NEMo package.  -->
<!-- from the online database [Atlas of the Living Australia](doi.ala.org.au/doi/10.26197/ala.d92678b1-ad2d-437b-9457-9f52737ba003) (see Appendix-method-occurence).  -->

## Statistical analyses
Statistical analyses were run in R version 4.2 (CITE R) and Julia (CITE). We visualised dynamics using baseR and `ggplot2` in R. The Australian bioregion map visualised in `QGIS` v (CITE Australian Map).

@@

## Biogeographic history

Based on our best fitting biogeographic model, the constrained-dispersal matrix DEC+_j_+_x_ model, we inferred tropical grassland biome to be ancestral biome of Australian _Anilios_ (\@ref(fig:fig-biogeobears). 
Likelihood ratio test (LRT) indicates that founder event (parameter *j*) and distance matrix (free parameter *x*) in the more complex model provides statistically significant improvement over the simpler DEC, DEC+*j*, and DEC+*x* models (X^2^, DF = 2, D-statistic = 26.7, p-value < 0.001). 
Our result also supports the hypothesis that the ancestor of the Australian clade first invaded the mesic part of the continent [@marinTracing2013], but exactly where they lived prior to this invasion is equivocal between the Bismarck Archipelago and New Guinea. 
After arriving to Australia in the late Oligocene - early Miocene (~25 MYA), the ancestral _Anilios_ initially diversified rapidly within less than 1 million years. Our analysis is consistent with the conclusion that Australian blindsnakes, like several other Australian squamates, originated from tropical-mesic origins and expanded into the arid zone since the mid-Miocene [@chappleParallel2004;@melvilleEvolutionary2011].

## Biome transitions and current distributions

Following the hypothesized colonisation event from a single common ancestor, the diversifcation of _Anilios_ occurred very rapidly. Descendents of the earliest split remained in the Tropical grassland biome but within a million years later another split occured. The earliest biome transition was from tropical grassland into Temperate Forest around 25 MYA (95% HPD: 28 – 22 MYA) and the Arid zone 24 MYA (95% HPD: 28 – 21 MYA). Subsequent splits show transitions from the Arid zone into the Mediterranean biome starting after 15 MYA (\@ref(fig:fig-biogeobears). Current distributions of _Anilios_ are mostly concentrated in the tropical grassland biome closely followed by the Arid zone and Mediterranean biome. We observed that closely related species tend to share the same biome and occur around the same geographical areas.

@@

Paleoclimate reconstructions of aridity differed slightly between the two methods \@ref(fig:fig-paleoclimate-compare)a. Both aridity index shows increase in aridity post 10 MYA, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021]. Paleoclimate estimates for Australia differed between the three methods \@ref(fig:fig-paleoclimate-compare)b. However, all three paleo temperature models indicates that Australia was hotter than the average global temperature. Paleoclimate in Australia were also relatively constant compared to global temperature that shows greater fluctuation then slight decrease towards the present.

## Environment-dependent diversification

We found that the exponential environment-dependent speciation and constant extinction (BEnvDCST) with aridity index derived from @scoteseAtlas2021 provided the best fit to the phylogeny (AICc = 238.43; see Table-RPANDA). This model indicates that increase aridity is exponentially associate with decrease speciation rate (lambda = 1.32, alpha = -0.14) and with a constant background extinction (mu = 0.23). The second and third most competitive models (delta AICc < 2) also found increase aridity is exponentially associated with decrease speciation rate but without estimating extinction (Table-RPANDA-TOP10). Among exponential environment-dependent models fitted to our blind snake phylogeny, we found that models that uses Australia-specific paleoclimate data received better AICc support compared to fit with global paleoclimate (see supp-TABLE_FULL_RPANDA_model_fit).

## Rates of diversification

@@

## Ancestral biome reconstruction
- Our result — that the ancestor of Anilios occupied tropical grasslands – is consistent with the result presented in @marinTracing2013. 
- We found transitions from XXX to XXX to be common as have been shown / as opposed to other Australian squamates (CITATIONS)
- Arid zone was not arid until ~20 Million Years ago. Ancestors in deeper time probably occupied temperate or tropical areas that subsequently became the arid zone we know see today. 
- The Australo-Papuan clade seems to have stayed in Australia for the most part but we were not able to include New Guinean endemic _Anilios erycinus_, the only recognised species of _Anilios_ not found in Australia. Therefore, we cannot make any conclusions about transitions between Australia and New Guinea even though at least three species (_A. torresianus_, _A. leucoproctus_, ...) are found in Northern 
Australia Torres Strait Islands, and southern parts of New Guinea.
- Range expansion was important for diversification ...
- Topography dyanmics [@brazModelling2021]
- Solomon Island 40 MYA [@hallCenozoic2002]

- Recent biogeographic reconstruction recovered India as the origin for superfamily Typhlopoidea (Gerrhopilidae, Xenotyphlopidae and Typhlopidae) [@kleinEvolution2021].
